---
title: "Grammar of Graphics and ggplot2"
subtitle: "SISBID 2016"
author: "Di Cook (dicook@monash.edu, @visnut); Heike Hofmann (heike.hofmann@gmail.com, @heike_hh)i; edited by Alexej Gossmann"
date: "`r Sys.Date()`"
output: html_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```

```{r echo=FALSE}
#install.packages(c("ggplot2", "tidyr", "dplyr", "lubridate",
#                   "scales", "readr", "ggmap", "HLMdiag", 
#                   "RColorBrewer", "gridExtra", "dichromat", "xkcd"))
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
library(scales)
library(readr)
library(ggmap)
library(HLMdiag)
library(RColorBrewer)
library(gridExtra)
library(dichromat)
library(xkcd)
```

## Using the package ggplot2

Elements of a plot

- data
- aesthetics: mapping of variables to graphical elements
- geom: type of plot structure to use
- transformations: log scale, ...

Additional components 

- layers: multiple geoms, multiple data sets, annotation
- facets: show subsets in different plots
- themes: modifying style

## Why use a grammar of graphics?

<font face="DidyBold" color="#ED9121" size=8> Variable in the data is directly mapped to an element in the plot </font>

## Data - Autism

```{r}
glimpse(autism)
```

## Plotting points

```{r}
ggplot(autism, aes(x=age2, y=vsae)) + 
  geom_point()
```

## Your turn

How is the data mapped to graphical elements?

- data: autism
- aesthetics: x and y positions along a line
- geom: point
- transformations: none

## Jittering points

```{r}
ggplot(autism, aes(x=age2, y=vsae)) + 
  geom_jitter()
```

## Your turn

How is the data mapped to graphical elements?

- data: autism
- aesthetics: x and y positions along a line
- geom: point
- transformations: jitter

## Adding lines

```{r}
ggplot(autism, aes(x=age2, y=vsae)) + 
  geom_point() + geom_line()
```

## Not the lines we want

```{r}
ggplot(autism, aes(x=age2, y=vsae, group=childid)) + 
  geom_point() + geom_line()
```

## Too much ink

```{r}
ggplot(autism, aes(x=age2, y=vsae, group=childid)) + 
  geom_point() + geom_line(alpha=0.5)
```

```{r}
ggplot(autism, aes(x=age2, y=vsae, group=childid)) + 
  geom_line(alpha=0.2) + theme_bw()
```
Now we can see that some individuals degrade, while most improve with time.

## Log scale y?

```{r}
ggplot(autism, aes(x=age2, y=vsae, group=childid)) + 
  geom_line(alpha=0.2) + scale_y_log10() + theme_bw()
```

## By age 2 diagnosis

```{r}
ggplot(autism, aes(x=age2, y=vsae, group=childid, colour=bestest2)) + 
  geom_line(alpha=0.3) + scale_y_log10() + theme_bw()
```
Now we can see a lot of overlap between the two groups.

## Refine groups

```{r}
ggplot(autism, aes(x=age2, y=vsae, colour=bestest2)) + 
  geom_point(alpha=0.1) + geom_line(aes(group=childid), alpha=0.1) + 
  geom_smooth(se=F) + scale_y_log10() + theme_bw()
```

```{r}
ggplot(autism, aes(x=age2, y=vsae, colour=bestest2)) + 
  geom_point(alpha=0.1) + geom_line(aes(group=childid), alpha=0.1) + 
  geom_smooth(se=F, method=lm) + scale_y_log10() + theme_bw()
```

## Your turn

What do we learn about autism, age, and the diagnosis at age 2?

>In terms of categorisation into either pdd or autism the vsae score is not distinct, but on average the autism diagnosis 2 year olds have lower scores. There a lot of overlap between ths group.

## Your turn

How is the data mapped to graphical elements?

- data: autism
- aesthetics: age on x and vsae on y, bestest2 mapped to color, smoothed vsae mapped to y
- geom: point, line
- transformations: log10 on y

## A different look

```{r}
ggplot(autism, aes(x=age2, y=vsae, colour=bestest2)) + 
  geom_boxplot() + scale_y_log10()
```

That's not what I wanted ....

## For each age measured

```{r}
ggplot(autism, aes(x=factor(age2), y=vsae, colour=bestest2)) + 
  geom_boxplot() + scale_y_log10()
```

## Which is better?

```{r}
p1 <- ggplot(autism, aes(x=age2, y=vsae, colour=bestest2)) + 
  geom_point(alpha=0.1) + geom_line(aes(group=childid), alpha=0.1) + 
  geom_smooth(se=F) + scale_y_log10() + theme(legend.position="none")
p2 <- ggplot(autism, aes(x=factor(age2), y=vsae, colour=bestest2)) + 
  geom_boxplot() + scale_y_log10() + theme(legend.position="none")
grid.arrange(p1, p2, ncol=2)
```

## New example - Flying etiquette

[41% Of Fliers Think You’re Rude If You Recline Your Seat](http://fivethirtyeight.com/datalab/airplane-etiquette-recline-seat/)

```{r}
fly <- read_csv("./data/flying-etiquette.csv")
glimpse(fly)
```

## Variables

Mix of categorical and quantiative variables. What mappings are appropriate? Area for counts of categories, side-by-side boxplots for mixed pair. 

## Support

```{r}
ggplot(fly, aes(x=`How often do you travel by plane?`)) + 
  geom_bar() + coord_flip()
```

Categories are not sorted

## Sorted categories

```{r}
fly$`How often do you travel by plane?` <- 
  factor(fly$`How often do you travel by plane?`, levels=c(
    "Never","Once a year or less","Once a month or less",
    "A few times per month","A few times per week","Every day"))
ggplot(fly, aes(x=`How often do you travel by plane?`)) +
  geom_bar() + coord_flip()
```

## Filter data

```{r}
fly_sub <- fly %>% filter(`How often do you travel by plane?` %in% 
                            c("Once a year or less",
                              "Once a month or less")) %>%
  filter(!is.na(`Do you ever recline your seat when you fly?`)) %>%
  filter(!is.na(Age)) %>% filter(!is.na(Gender))
```

## Recline by height

```{r}
fly_sub$`Do you ever recline your seat when you fly?` %>% unique()

fly_sub$`Do you ever recline your seat when you fly?` <- factor(
  fly_sub$`Do you ever recline your seat when you fly?`, levels=c(
    "Never","Once in a while","About half the time",
    "Usually","Always"))
ggplot(fly_sub, aes(y=`How tall are you?`, 
                    x=`Do you ever recline your seat when you fly?`)) + 
                      geom_boxplot() #+ coord_flip()
```

## Cheat sheet

Take a look at the ggplot2 [Cheat sheet](http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf)

## Your turn

How many geoms are available in ggplot2? What is `geom_rug`?

```{r}
p <- ggplot(autism, aes(x=age2, y=vsae))
p1 <- p + geom_point() + coord_flip()
p2 <- p + geom_point() + geom_rug() + coord_flip()
p3 <- p + geom_point() + geom_rug(position='jitter') + coord_flip()
grid.arrange(p1, p2, p3, nrow=3)
```

## Your turn

What is the difference between `colour` and `fill`?

*Colour is for 0 or 1-dimensional elements, and fill is for area (2-d) geoms*

## Your turn

What does `coord_fixed()` do? What is the difference between this and using `theme(aspect.ratio=...)`?

```{r}
p <- ggplot(autism, aes(x=age2, y=vsae))
p1 <- p + geom_point() + coord_fixed(ratio = 1)
p2 <- p + geom_point() + theme(aspect.ratio = 1)
grid.arrange(p1, p2, ncol=2)
```

`coord_fixed` operates on the raw data values, but `theme(aspect_ratio=...)` works on the plot dimensions.

## Your turn

What are scales? How many numeric transformation scales are there?

scales do the transformation between data values and graphical element value. most often it is applied to position along x, y which is common, to log or sqrt, .. there are 3 numeric transformations. 

## Your turn

What are position adjustments? When would they be used?

positions shift the location some from original coordinates. most often used with bar charts to stack, or put side-by-side

## Your turn 

Use your cheat sheet to work out how to make a plot to explore the relationship between 

`Do you ever recline your seat when you fly?` and `Is it rude to recline your seat on a plane?`


```{r}
unique(fly_sub$`Is itrude to recline your seat on a plane?`)
unique(fly_sub$`Do you ever recline your seat when you fly?`)
ggplot(fly_sub, aes(x=`Do you ever recline your seat when you fly?`)) +
  geom_bar() + 
  facet_wrap(~`Is itrude to recline your seat on a plane?`, ncol=3) +
  coord_flip()
```

```{r}
ggplot(fly_sub, aes(x=`Do you ever recline your seat when you fly?`,
                    fill=`Is itrude to recline your seat on a plane?`)) +
  geom_bar()
```

```{r}
ggplot(fly_sub, aes(x=`Do you ever recline your seat when you fly?`,
                    fill=`Is itrude to recline your seat on a plane?`)) +
  geom_bar(position="dodge")
```

## Facets

```{r}
ggplot(fly_sub, 
       aes(x=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar() + coord_flip() + facet_wrap(~Gender)
```

## Facets

```{r}
fly_sub$Age <- factor(fly_sub$Age, 
                      levels=c("18-29","30-44","45-60","> 60"))
ggplot(fly_sub, 
       aes(x=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar() + coord_flip() + facet_grid(Age~Gender)
```

## Color palettes - default

```{r}
p <- ggplot(fly_sub, 
            aes(x=`In general, is itrude to bring a baby on a plane?`,
                fill=Gender)) + 
  geom_bar(position="fill") + coord_flip() + 
  facet_wrap(~Age, ncol=5)
p
```

## Color palettes - brewer

```{r}
p + scale_fill_brewer(palette="Dark2") 
```

## Color blind-proofing

What it looks like to a color-blind:

```{r}
library(scales)
library(dichromat)
p1 <- p + theme(legend.position = "none")
clrs <- hue_pal()(3)
clrs <- dichromat(clrs)
p2 <- p + scale_fill_manual("", values=clrs) + 
  theme(legend.position = "none")
grid.arrange(p1, p2)
```


## Perceptual principles

- Hierarchy of mappings: (first) position along an axis - (last) color (Cleveland, 1984; Heer and Bostock, 2009)
- Pre-attentive: Some elements are noticed before you even realise it.
- Color: (pre-attentive) palettes - qualitative, sequential, diverging.
- Proximity: Place elements for primary comparison close together. 
- Change blindness: When focus is interrupted differences may not be noticed.

## Hierarchy of mappings

1. Position - common scale (BEST)
2. Position - nonaligned scale
3. Length, direction, angle
4. Area
5. Volume, curvature
6. Shading, color (WORST)

## Pre-attentive

Can you find the odd one out?

```{r}
df <- data.frame(x=runif(100), y=runif(100), 
                 cl=sample(c(rep("A", 1), rep("B", 99))))
ggplot(data=df, aes(x, y, shape=cl)) + theme_bw() + 
  geom_point() + theme(legend.position="None", aspect.ratio=1)
```

---

Is it easier now?

```{r}
ggplot(data=df, aes(x, y, colour=cl)) + 
  geom_point() + theme_bw() + 
  theme(legend.position="None", aspect.ratio=1)
```


## Color palettes

- Qualitative: categorical variables
- Sequential: low to high numeric values
- Diverging: negative to positive values

---

```{r}
library(RColorBrewer)
display.brewer.all()
```


## Proximity

```{r}
ggplot(fly_sub, aes(x=`In general, is itrude to bring a baby on a plane?`,
                    fill=Gender)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5)
```

With this arrangement we can see proportion of gender within each rudeness category, and compare these across age groups.  How could we arrange this differently?


## Proximity

```{r}
ggplot(fly_sub, aes(x=Gender, 
                    fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5) +
  theme(legend.position="bottom")
```

##

```{r}
ggplot(fly_sub, aes(x=Gender,
                    fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5) + theme(legend.position="bottom")
```

What is different about the comparison now?

## Another arrangement

```{r}
ggplot(fly_sub, aes(x=Age,
                    fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Gender, ncol=5) + 
  theme(legend.position="bottom")
```

## Themes

The `ggthemes` package has many different styles for the plots. Other packages such as `xkcd`, `skittles`, `wes anderson`, `beyonce`, ....

```{r}
ggplot(fly_sub, aes(x=Gender,
                    fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5) +
  theme_xkcd() + theme(legend.position="bottom")
```

See the [vignette](https://cran.r-project.org/web/packages/xkcd/vignettes/xkcd-intro.pdf) for instructions on installing the xkcd font. 

##

```{r}
ggplot(fly_sub, aes(x=Gender,
                    fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5) +
  theme_xkcd() + theme(legend.position="bottom")
```

## Resources

- [Cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf)
- [ggplot2: Elegant Graphics for Data Analysis, Hadley Wickham](http://ggplot2.org/book/), [web site](http://ggplot2.org)
- [R Graphics Cookbook, Winston Chang](http://www.cookbook-r.com/Graphs/)
- [Naomi Robbins, Creating More Effective Graphs](http://www.nbr-graphs.com)
- [Antony Unwin, Graphical Data Analysis with R](https://www.crcpress.com/Graphical-Data-Analysis-with-R/Unwin/9781498715232)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
