---
title: "Multivariate data plots"
subtitle: "SISBID 2016"
author: "Di Cook (dicook@monash.edu, @visnut); Heike Hofmann (heike.hofmann@gmail.com, @heike_hh)i; edited by Alexej Gossmann"
date: "`r Sys.Date()`"
output: html_document
---

```{r echo = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```

```{r}
library(readr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(GGally)
library(edgeR)
```

## Main types of plots

- pairwise plots: explore replicates, and associations
- porcupine plots: to explore treatments vs replicates
- parallel coordinate plots: explore high-dimensional relationships with parallel axes

## GGally

Classical scatterplot matrix (draftman's) plot. All variables are numeric, show all possible pairs.


```{r}
data(flea)
ggscatmat(flea, columns = 2:4, color = "species") 
```

##

Generalized pairs plot, any combination of types of plots

```{r fig.width=6}
data(tips, package = "reshape")
ggpairs(tips[, 1:4])
```

## For biological data

Brown, A. and Hudson, K. (2015) Developmental profiling of gene expression in soybean trifoliate leaves and cotyledons, [GSE61857](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE61857). 

```{r fig.width=6, fig.height=6}
coty <- read_delim("data/GSE61857_Cotyledon_normalized.txt.gz",
  delim="\t", col_types="cddddddddd", 
  col_names=c("ID", "C_S1_R1", "C_S1_R2", "C_S1_R3", 
  "C_S2_R1", "C_S2_R2", "C_S2_R3", "C_S3_R1", "C_S3_R2", "C_S3_R3"),
  skip=1)
coty <- as.data.frame(coty)
glimpse(coty)
ggscatmat(coty, columns=2:7, alpha=0.1)
```

Normalized data for the first 6 samples are shown.

## What we learn

- The replicates look more like each other than the developmental stage 1 vs 2. This is great data!
- There is a lot of variability in counts across stages, would indicate a lot of gene activity. 
- There are a few genes that have some difference from one rep to another. These could be problematic when it comes to testing. To identify these it is easier to use interactive graphics - stay tuned for this tomorrow.

## Your turn

- Read the leaf data in from Brown and Hudson, and make a scatterplot matrix. (If it is too slow on your computer, sample the number of rows, to get a glimpse of the data to start.)
- Summarise what you learn about the data in a couple of sentences

```{r}
leaf <- read_delim("data/GSE61857_Leaf_normalized.txt.gz",
                   delim="\t", col_types="cddddddddddddddd", 
                   col_names=c("ID", "L_S1_R1", "L_S1_R2", "L_S1_R3", 
                               "L_S2_R1", "L_S2_R2", "L_S2_R3", 
                               "L_S3_R1", "L_S3_R2", "L_S3_R3",
                               "L_S4_R1", "L_S4_R2", "L_S4_R3", 
                               "L_S5_R1", "L_S5_R2", "L_S5_R3"),
                   skip=1)
leaf <- as.data.frame(leaf)
ggscatmat(leaf, columns=2:7, alpha=0.1)
```

## Porcupine plots

Compare treatment vs reps. Focus on reps 1, 2 from stages 1 and 2. 

```{r}
sub <- coty %>% select(ID, C_S1_R1, C_S1_R2, C_S2_R1, C_S2_R2)
ggplot(sub, aes(x=C_S1_R1, xend=C_S1_R2, y=C_S2_R1, yend=C_S2_R2)) +
  geom_segment() + 
  theme(aspect.ratio = 1)
```

The best prospects should be shortest lines further from the `X=Y`. 

## Your turn

Make a porcupine plot for the first two reps of the first two developmental stages of the leaf data.

```{r}
names(leaf)
leaf.sub <- leaf %>% select(ID, L_S1_R1, L_S1_R2, L_S2_R1, L_S2_R2)
ggplot(leaf.sub, aes(x=L_S1_R1, xend=L_S1_R2, y=L_S2_R1, yend=L_S2_R2)) +
  geom_segment() + 
  theme(aspect.ratio = 1)
```
- Not as clean as the other data, because there are not as many short lines in the extremes (we want small rep differences and large treatment differences).
- Many horizontal lines may imply some problem with the normalization between reps.

## Parallel coordinate plots 

- Parallel coordinate plots, use parallel axes, and each observation is represented by a line tracing through the axes.
- They are useful for getting a sense of the multivariate shape of the data. You can typically read association, clustering and outliers from the plot. 
- Can handle more variables than scatterplot matrices.

## Side-by-side boxplots to par coords

```{r}
ggparcoord(coty, columns=2:10, scale="globalminmax", boxplot=TRUE,
           alphaLines=0)
```

If the data is normalised well these should be identical boxplots, or very close to that. This is not quite what we see. 

## Parallel coordinates, connect the dots


```{r eval=F}
ggparcoord(coty, columns=2:10, scale="globalminmax", 
           alphaLines=0.1)
```

```{r}
ggparcoord(coty[sample(1:nrow(coty), 1000),], 
           columns=2:10, scale="globalminmax", 
           alphaLines=0.1)
```

## Scaling choices

- *std*: univariately, subtract mean and divide by standard deviation
- *robust*: univariately, subtract median and divide by median absolute deviation
- *uniminmax*: univariately, scale so the minimum of the variable is zero, and the maximum is one
- *globalminmax*: no scaling is done; the range of the graphs is defined by the global minimum and the global maximum

## Your turn

- Make parallel coordinate plots for the 9 samples of the leaf data
- Start with side-by-side boxplots, and then generate the lines
- Run the different scaling options to discover how they change the appearance of the plot

```{r}
leaf.sub <- leaf[sample(1:nrow(leaf), 1000), ]
ggparcoord(leaf.sub, columns=2:10, scale="globalminmax", boxplot=TRUE,
           alphaLines=0)
```

```{r}
ggparcoord(leaf.sub, 
           columns=2:10, scale="globalminmax", 
           alphaLines=0.1)
```

## Parcoords to interaction plots

Experiments are often displayed using interaction plots. These are similar to parallel coordinates, which display all the measurements, but interaction plots show one gene at a time.

##

Process the data, and then plot

```{r}
sub <- coty[sample(1:nrow(coty), 6),] %>%
  gather(sample, expr, -ID) %>%
  separate(sample, c("tissue", "stage", "rep")) %>%
  mutate(stage = as.numeric(substr(stage, 2, 2)))
sub.m <- sub %>% group_by(ID, stage) %>% 
  summarise(expr = mean(expr))
```

##


```{r}
ggplot(sub, aes(x=stage, y=expr)) + geom_point(alpha=0.4) +
  facet_wrap(~ID, ncol=3) + 
  geom_line(data=sub.m, aes(x=stage, y=expr), colour="red")
```

## Your turn

Re-make the previous plot, by overlaying them in one plot, instead of faceting, and colour by the ID. 

```{r}
ggplot(sub, aes(x=stage, y=expr, color=factor(ID), shape=factor(ID))) + 
  geom_jitter(alpha=0.4) +
  geom_line(data=sub.m, aes(x=stage, y=expr))
```

## With significance testing

```{r}
d <- DGEList(counts = coty[,-1], 
  group = c(rep("S1", 3), rep("S2", 3), rep("S3", 3)), 
  genes = coty[,1])
d <- calcNormFactors(d)
d <- estimateCommonDisp(d)
d <- estimateTagwiseDisp(d)
de <- glmFit(d, design=matrix(c(rep(1, 9),
                                c(1,1,1,0,0,0,0,0,0), 
                                c(0,0,0,1,1,1,0,0,0)),
                              ncol=3, byrow=F))
results <- glmLRT(de)
top10 <- as.data.frame(topTags(results, n=10))
```

## Plot

```{r}
sub <- coty %>% 
  gather(sample, expr, -ID) %>%
  separate(sample, c("tissue", "stage", "rep")) %>%
  mutate(stage = as.numeric(substr(stage, 2, 2)))
top10 <- merge(top10, sub, by.x = "genes", by.y = "ID")
top10.m <- top10 %>% group_by(genes, stage) %>% 
  summarise(expr = mean(expr))
ggplot(top10, aes(x=stage, y=expr, colour=genes)) + 
  geom_point(alpha=0.4) + ylim(c(0,17.5)) +
  geom_line(data=top10.m, aes(x=stage, y=expr, colour=genes))
```

## Your turn

Plot the next 10 most significant genes. 

```{r}
top10 <- as.data.frame(topTags(results, n=24))[13:24,]
top10 <- merge(top10, sub, by.x = "genes", by.y = "ID")
top10.m <- top10 %>% group_by(genes, stage) %>% 
  summarise(expr = mean(expr))
ggplot(top10, aes(x=stage, y=expr, colour=genes)) + 
  geom_point(alpha=0.4) + ylim(c(0,17.5)) +
  geom_line(data=top10.m, aes(x=stage, y=expr, colour=genes))
```

## Extracting the significant genes

Just look pairwise for simplicity here.

```{r}
d <- DGEList(counts = coty[,2:7], 
  group = c(rep("S1", 3), rep("S2", 3)), 
  genes = coty[,1])
d <- calcNormFactors(d)
d <- estimateCommonDisp(d)
d <- estimateTagwiseDisp(d)
d <- estimateTrendedDisp(d)
de <- exactTest(d, pair=c("S1", "S2"), dispersion = "trended")
sig.tab <- de$table
sig.tab$genes <- coty$ID
sig.tab <- sig.tab %>% filter(PValue < 0.01)
```

##

```{r}
head(sig.tab)
head(coty[,1:7])
sig.tab <- merge(sig.tab, coty[,1:7], by.x="genes", by.y="ID")
ggscatmat(sig.tab, columns=5:10, alpha=0.1)
```

##

```{r}
ggparcoord(sig.tab, columns=5:10, scale="globalminmax",
           alphaLines=0.1)
```

## Resources

- [GGobi web site](http://www.ggobi.org), [ggobi book](http://www.ggobi.org/book)
- Emerson et al (2013) The Generalized Pairs Plot, Journal of Computational and Graphical Statistics, 22:1, 79-91
- Cook et al (2007) Exploring Gene Expression Data, Using Plots, {\em Journal of Data Science, 5(2):151-182
- [Antony Unwin, Graphical Data Analysis with R](https://www.crcpress.com/Graphical-Data-Analysis-with-R/Unwin/9781498715232)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
