---
title: "Tidying your data"
subtitle: "SISBID 2016"
author: "Di Cook (dicook@monash.edu, @visnut); Heike Hofmann (heike.hofmann@gmail.com, @heike_hh); edited by Alexej Gossmann"
date: "`r Sys.Date()`"
output: html_document
---

```{r echo = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```

```{r}
library(knitr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
library(scales)
library(readr)
library(ggmap)
library(HLMdiag)
library(RColorBrewer)
library(gridExtra)
library(broom)
```

## Using the Packages `tidyr`, `dplyr`

- Writing readable code using **pipes**
- What is **tidy data**? Why do you want tidy data? Getting your data into tidy form using tidyr.
- Reading different **data formats**
- String operations, working with **text**

## Your turn 1

What are the variables?

```{r}
grad <- read_csv("./data/graduate-programs.csv")
kable(head(grad[c(2,3,4,6)]))
```
in the columns, subject, Inst, AvNumPubs, ...

## Your turn 2

What's in the column names of this data? What are the experimental units? What are the measured variables?

```{r}
genes <- read_csv("./data/genes.csv")
kable(genes)
```

the experimental design is coded into the variable names, genotype:WI/WM, time:6/12, rep:1/2/4

## Your turn 3

What are the variables? What are the records?

```{r}
melbtemp <- read.fwf("./data/ASN00086282.dly", 
   c(11, 4, 2, 4, rep(c(5, 1, 1, 1), 31)), fill=T)
kable(head(melbtemp[,c(1,2,3,4,seq(5,128,4))]))
```

variables are TMAX, TMIN, PRCP, year, month, day, stationid. Each row contains the values for one month!

## Your turn 4

What are the variables? What are the experimental units?

```{r}
tb <- read_csv("./data/tb.csv")
tail(tb)
colnames(tb)
```

## Your turn 5 

What are the variables? What are the experimental units?

```{r}
pew <- read.delim(
  file = "http://stat405.had.co.nz/data/pew.txt",
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = F
)
kable(pew[1:5, 1:5])
```

## Your turn 6

10 week sensory experiment, 12 individuals assessed taste of french fries on several scales (how potato-y, buttery, grassy, rancid, paint-y do they taste?), fried in one of 3 different oils, replicated twice. First few rows:

```{r}
data(french_fries, package = "reshape2")
kable(head(french_fries, 4), format = "markdown", row.names = F)
```

What is the experimental unit? What are the factors of the experiment? What was measured? What do you want to know?

## Messy Data Patterns

There are various features of messy data that one can observe in practice. Here are some of the more commonly observed patterns.

- Column headers are values, not variable names
- Variables are stored in both rows and columns, contingency table format
- One type of experimental unit stored in multiple tables
- Dates in many different formats

## What is Tidy Data?

- Each observation forms a row
- Each variable forms a column
- Contained in a single table
- Long form makes it easier to reshape in many different ways
- Wide form is common for analysis

## Tidy Verbs

- **gather**: specify the **keys** (identifiers) and the **values** (measures) to make long form (used to be called melting)
- **spread**: variables in columns (used to be called casting)
- **nest/unnest**: working with lists
- **separate/unite**: split and combine columns

## French Fries 

10 week sensory experiment, 12 individuals assessed taste of french fries on several scales (how potato-y, buttery, grassy, rancid, paint-y do they taste?), fried in one of 3 different oils, replicated twice. First few rows:

```{r, echo=FALSE, results='asis'}
kable(head(french_fries), format = "markdown", row.names = FALSE)
```

## What Would We Like to Know?

- Is the design complete?
- Are replicates like each other?
- How do the ratings on the different scales differ?
- Are raters giving different scores on average?
- Do ratings change over the weeks?

Each of these questions involves different summaries of the data.

## Gathering

- When gathering, you need to specify the **keys** (identifiers) and the **values** (measures).

Keys/Identifiers:
- Identify a record (must be unique)
- Example: Indices on an random variable
- Fixed by design of experiment (known in advance)
- May be single or composite (may have one or more variables)

Values/Measures:
- Collected during the experiment (not known in advance)
- Usually numeric quantities

## Gathering the French Fries

```{r}
ff_long <- gather(french_fries, key = variable, value = rating, 
                  potato:painty)
head(ff_long)
```

## Long to Wide

In certain applications, we may wish to take a long dataset and convert it to a wide dataset (perhaps displaying in a table).

This is called "spreading" the data.

## Spread

We use the **spread** function from tidyr to do this:

```{r}
french_fries_wide <- spread(ff_long, key = variable, 
                            value = rating)
head(french_fries_wide)
```

## Answer some Questions

- Easiest question to start is whether the ratings are similar on the different scales, potato'y, buttery, grassy, rancid and painty. 

- We need to gather the data into long form, and make plots facetted by the scale. 

## Ratings on the Different Scales

```{r}
ff.m <- french_fries %>% 
  gather(type, rating, -subject, -time, -treatment, -rep)
head(ff.m)
```

##

```{r, fig.height=2, fig.width=8}
ggplot(data=ff.m, aes(x=rating)) + geom_histogram(binwidth=2) + 
  facet_wrap(~type, ncol=5) 
```

## Side-By-Side Boxplots

```{r fig.width=8, fig.height=5}
ggplot(data=ff.m, aes(x=type, y=rating, fill=type)) + 
  geom_boxplot()
```


## Do the Replicates Look Like Each Other?

We will start to tackle this by plotting the replicates against each other using a scatterplot. 

We need to gather the data into long form, and then get the replicates spread into separate columns. 

## Check Replicates

```{r}
head(ff.m)
ff.s <- ff.m %>% spread(rep, rating)
head(ff.s)
```

## Check Replicates

```{r, fig.show='hold', fig.align='default', fig.height=4, fig.width=4}
ggplot(data=ff.s, aes(x=`1`, y=`2`)) + geom_point() +
  theme(aspect.ratio=1) + xlab("Rep 1") + ylab("Rep 2")
ggplot(data=ff.s, aes(x=`1`, y=`2`, color=type)) + geom_point() +
  theme(aspect.ratio=1) + xlab("Rep 1") + ylab("Rep 2") + 
  scale_x_sqrt() + scale_y_sqrt()
```

## Your turn

Make the scatterplots of reps against each other separately for scales, and treatment. 

```{r}
ggplot(data=ff.s, aes(x=`1`, y=`2`)) + geom_point() +
  theme(aspect.ratio=1) + 
  scale_x_sqrt() + scale_y_sqrt() +
  xlab("Rep 1") + ylab("Rep 2") + facet_wrap(~type, nrow=2, ncol=3)
ggplot(data=ff.s, aes(x=`1`, y=`2`)) + geom_point() +
  theme(aspect.ratio=1) + 
  scale_x_sqrt() + scale_y_sqrt() +
  xlab("Rep 1") + ylab("Rep 2") + facet_grid(treatment~type)
```

## Your turn

Read in the billboard top 100 music data, which contains N'Sync and Backstreet Boys songs that entered the billboard charts in the year 2000

```{r}
billboard <- read.csv("./data/billboard.csv")
dim(billboard)
colnames(billboard)
billboard[ , 1:10]
billboard[ , 30:41]
```

What's in this data? What's X1-X76?

position in the billboard top 100

## Your turn

1. Use `tidyr` to convert this data into a long format appropriate for plotting a time series (date on the x axis, chart position on the y axis)
2. Use `ggplot2` to create this time series plot:

```{r}
lf <- billboard %>% gather(week, position, -year, -artist, -track,
                           -time, -date.entered)
lf %>% head
lf$week <- as.numeric(gsub("X", "", lf$week))
lf <- filter(lf, !is.na(position))
ggplot(data = lf, aes(x=week, y=position, colour = artist, group = track)) + geom_line() + theme(legend.position="bottom")
```

## String Manipulation

When the experimental design is packed into column names, we need to extract it, and tidy it up. 

```{r}
genes <- read_csv("./data/genes.csv")
head(genes)
```

## Gather Column Names into Long Form

```{r}
gather(genes, variable, expr, -id)
```

## Separate Columns

```{r}
genes %>%
  gather(variable, expr, -id) %>%
  separate(variable, c("trt", "leftover"), "-") 
```

##

```{r}
genes %>%
  gather(variable, expr, -id) %>%
  separate(variable, c("trt", "leftover"), "-") %>%
  separate(leftover, c("time", "rep"), "\\.")
```

##

```{r}
gtidy <- genes %>%
  gather(variable, expr, -id) %>%
  separate(variable, c("trt", "leftover"), "-") %>%
  separate(leftover, c("time", "rep"), "\\.") %>%
  mutate(trt = gsub("W", "", trt)) %>%
  mutate(rep = as.numeric(gsub("R", "", rep)))
head(gtidy)
```

## Your turn

(1) Using the tidied dataset (`gtidy`), find the mean expression for each combination of id, trt, and time.
(2) Use this tidied data to make this plot.

```{r}
mean.expr <- gtidy %>% group_by(id, trt, time) %>%
  summarize(mean.expr=mean(expr))
head(mean.expr)

ggplot(mean.expr, aes(x=trt, y=mean.expr, color=time)) +
  geom_point() + facet_wrap(~id) +
  xlab("Type of modification") + ylab("Expression") + 
  geom_line(data=mean.expr, aes(group = time))
```

## Tidying model output

The package `broom` gets model results into a tidy format. 

```{r}
ff_long <- gather(french_fries, key = variable, value = 
                    rating, potato:painty)
head(ff_long, 2)
ff_lm <- lm(rating~variable+treatment+time+subject, 
            data=ff_long)
ff_lm %>% summary
```

## Model estimates

```{r}
ff_lm_tidy <- tidy(ff_lm)
glimpse(ff_lm_tidy)
```

## Model diagnostics

```{r}
ff_lm_all <- augment(ff_lm)
glimpse(ff_lm_all)
```

## Residual plot

```{r}
ggplot(ff_lm_all, aes(x=.fitted, y=.resid)) + geom_point()
```

## Back to autism

Fit a random slopes model to the data:

$$vsae = \beta_0 + \beta_1 age2 +  b_{childid} age2 + \varepsilon$$

```{r}
library(lme4)
data(autism)
head(autism, 3)
autism_keep <- autism %>% group_by(childid) %>% 
  tally(sort=TRUE) %>% filter(n>2)
autism_sub <- autism %>% 
  filter(childid %in% autism_keep$childid)
autism_lmer <- lmer(vsae ~ age2 + ( age2 - 1 | childid ), 
                    data = autism_sub)
summary(autism_lmer)
```

## Your turn

- Augment the autism data with the model diagnostics
- Plot the residuals from the (a) fixed, (b) random effects
- Color the residuals by (a) gender, (b) besttest2
- Plot fitted values against observed values

```{r}
autism_lmer_all <- augment(autism_lmer, autism_sub)
glimpse(autism_lmer_all)
ggplot(autism_lmer_all, aes(x=.fitted, y=.fixed)) + geom_point()
ggplot(autism_lmer_all, aes(x=.fitted, y=.resid)) + geom_point()
ggplot(autism_lmer_all, aes(x=.fitted, y=.resid, colour=gender)) +
  geom_point()
ggplot(autism_lmer_all, aes(x=.fitted, y=.resid, colour=bestest2)) +
  geom_point()
ggplot(autism_lmer_all, aes(x=.fitted, y=vsae, label=childid)) +
  geom_point()
library(plotly)
ggplotly()
ggplot(autism_lmer_all, aes(x=.fitted, y=vsae, label=childid)) +
  geom_point() + geom_line(aes(group=childid))
```

## Resources

- [tidyr and dplyr cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
- [Wickham (2007) Reshaping data](https://www.jstatsoft.org/article/view/v021i12/v21i12.pdf)
- [broom vignettes, David Robinson](https://cran.r-project.org/web/packages/broom/vignettes/broom.html)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
